---
title: "Concord_MA_Use"
output: pdf_document
date: "2024-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
install.packages("tidyverse")
library(tidyverse)
```

## Read csv file 
```{r}
install.packages("readr")
library(readr)
Concord_data <- read_csv("Observations_of_flowering_phenology_in_Concord__Massachusetts__USA__1963-1993.csv")
```

## Pivot data from wide to long format 
```{r}
install.packages("dplyr")
library(dplyr)
Concord_long <- Concord_data %>% 
  pivot_longer(
    cols = 4:61,
    names_to = "year",
    values_to = "date")
```

## X is only removed in Posit Cloud - remove the X from the beginning of our year 

Seperate Pheno phase by start and end 
```{r}
Concord_long <- Concord_long %>%
  mutate(
    Pheno = 
      str_extract(year,"[a-z]+") )
```

Remove "s" and "e" from year
```{r}
Concord_long$year <-  
  str_remove(
    Concord_long$year, "[es]")
```

Extract the day from date column 
```{r}
Concord_long <- Concord_long %>%
  mutate(day  =
  str_extract(date, "\\-*\\d+\\.*\\d*")
)
```

Remove day  from date column to just leave year
```{r}
Concord_long$date <- 
      str_sub(Concord_long$date, 3)
```

Remove - from date then make dates into numbers not text
```{r}
Concord_long$date <- Concord_long$date %>% 
      str_replace_all("-","")

Concord_long <- Concord_long %>%
  rename(month = date)

month_mapping <- c(jan = "01", feb = "02", mar = "03", apr = "04", may = "05", jun = "06",
                   jul = "07", aug = "08", sep = "09", oct = "10", nov = "11", dec = "12")

Concord_long <- Concord_long %>%
  mutate(month = ifelse(!is.na(month), month_mapping, NA))
```

Create new date column using YMD format
```{r}
Concord_long <- Concord_long %>% 
  mutate(Observation_date = make_date(year = year, month = month, day = day))
```

Erase month, day, year column
```{r}
Concord_long <- Concord_long [-c(4,5,7)]
```

Add kingdom column
```{r}
Concord_long <- Concord_long %>% 
    mutate( Kingdom = "Plantae")

Concord_long <-  Concord_long %>% relocate(Kingdom)
```

Remove any rows that have NA in observation date column 
```{r}
Concord_long <- 
  Concord_long %>% 
    drop_na()
```

Replace S and E with Start and End
```{r}
Concord_long$Pheno <-  
  str_replace(Concord_long$Pheno, "s", "Start")

Concord_long$Pheno <-  
  str_replace(Concord_long$Pheno, "e", "End")
```

Add complete scientific name 
```{r}
Concord_long <- Concord_long %>%
  mutate(Scientific_name = paste(Concord_long$Genus,Concord_long$Species))

Concord_long$Scientific_name <- str_replace_all(Concord_long$Scientific_name, "\\s{2,}", " ")

Concord_long <-  Concord_long %>% relocate(Scientific_name,.after = Species)

```

add column for lat and long
```{r}
Concord_long <- Concord_long %>%
  mutate(Longitude = "-71.3691069")

Concord_long <- Concord_long %>%
  mutate(Latitude = "42.4403334")
```

attach tsn number - first create a test df that can be joined to concord long df. Test df will only include unique species name 
```{r}
install.packages("taxize")
library(taxize)

Concord_long$Scientific_name <- paste(Concord_long$Genus,Concord_long$Species)

test_df <- distinct(Concord_long,Scientific_name)

test_df <- test_df %>% 
  mutate(
    Scientific_name=
      str_squish(Scientific_name))

test_df <- test_df %>%
  mutate(TSN = get_tsn(test_df$Scientific_name, rows = 1))

```

Create new df that only has species where TSN = NA - prob dont need this
```{r}
NA_TSN <- test_df[is.na(test_df$TSN),]
```

Use fuzzymatch from worldflora to get the correct name for NA TSN in test_df -> will compare TSN from test_df and worldflora to make sure no one is different 
```{r}
install.packages("WorldFlora")
library(WorldFlora)

install.packages("fuzzyjoin")
library(fuzzyjoin)

WFO.download()

WFO.remember("classification.csv")

fuzzy_match <- WFO.match.fuzzyjoin(spec.data = NA_TSN$Scientific_name, WFO.data = WFO.data, fuzzydist.max = 2)

# have WFO condense the above dataframe down to one clear match per species
one_to_one <- WFO.one(fuzzy_match)

# save the output so we don't have to run those again because they take up wild amounts of RAM!
write_csv(one_to_one, "corrected_taxonomy.csv")

# run the updated names through `taxize` to get updates TSN
one_to_one_TSN <- one_to_one %>% 
  mutate(TSN = get_tsn(one_to_one$scientificName, rows = 1))

# add tsn to Silene pennsylvanica

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate( 
    TSN = ifelse(spec.name == "Silene pennsylvanica", "20057", TSN)) 

```

Rename the scientific_name column in one to one TSN to match test_df
```{r}
colnames(one_to_one_TSN)[colnames(one_to_one_TSN) == 'scientificName'] <- 'Scientific_name'

```

Change TSN from chr to numeric
Relocate positions of columns to match test_df
```{r}
test_df <- test_df %>% 
  mutate(TSN = as.numeric(test_df$TSN)) 

one_to_one_TSN <- one_to_one_TSN %>% 
  mutate(TSN = as.numeric(one_to_one_TSN$TSN)) 

one_to_one_TSN <-  one_to_one_TSN %>% 
  relocate(Scientific_name, .before = spec.name)

one_to_one_TSN <-  one_to_one_TSN %>% 
  relocate(TSN, .after = Scientific_name)
```

Rename all misspelled Scientific_name in test_df to correct name located in one_to_one_TSN
Replace all NA values in test_df with values from one_to_one_TSN
```{r}
### don't erase until you run the full program -- i'm trying to see if it actually works 
misspelled <- one_to_one_TSN$spec.name

for (i in 1:nrow(test_df)) {
  if (test_df$Scientific_name[i] %in% misspelled){
    index <- which(one_to_one_TSN$spec.name == test_df$Scientific_name [i])
    test_df$Scientific_name[i] <- one_to_one_TSN$Scientific_name[index]
  } else {
    test_df$Scientific_name[i] <- test_df$Scientific_name [i]
  }
}
### run this after top for loop is confirmed to work
for (i in 1:nrow(test_df)){
  if (is.na(test_df$TSN[i])){
    index <- which(one_to_one_TSN$Scientific_name == test_df$Scientific_name [i])
    test_df$TSN[i] <-  one_to_one_TSN$TSN[index]
  }else {
    test_df$TSN [i] == test_df$TSN [i]
  }
}
  


i=11
```

Fix all misspelled names in Concord_Long -- do we fix species and genus 
```{r}

Concord_corrected_name <- read.csv("corrected_taxonomy.csv")

misspelled <- Concord_corrected_name$spec.name
for (i in 1:nrow(Concord_long)) {
  if (str_trim(Concord_long$Scientific_name[i]) %in% misspelled){
    index <- which(Concord_corrected_name$spec.name.ORIG == str_trim(Concord_long$Scientific_name[i]))
    Concord_long$Scientific_name[i] <- Concord_corrected_name$scientificName[index]
  } else {
    Concord_long$Scientific_name[i] <- Concord_long$Scientific_name [i]
  }
}

```


Final step
```{r}
Clean_Concord_Data <- full_join(Concord_long,test_df)

```

``````

